---
title: "Tutorial: bulk RNA-seq deconvolution using BayesPrism"
author: "Sophia Zhao"
date: "2023-12-13"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Due to the computational ability, the code up to runing BayesPrism will not be run through. (We run excute_PBMC.R file in Compute Canada.)
Here we use PBMC data, the procedure of pancreas is the same, just changing the data input.

Load the BayesPrism package
```{r library, eval = FALSE}
# install.packages("snowfall")
# install.packages("NMF")
# install.packages("gplots")
# install.packages("scran")
# install.packages("BiocParallel")
# install.packages("devtools")
library(snowfall)
library(NMF)
library(gplots)
library(scran)
library(BiocParallel)
library(devtools)
# install_github("Danko-Lab/BayesPrism/BayesPrism")
suppressWarnings(library(BayesPrism))
```

## Load the data

`bk.dat`: The sample-by-gene raw count matrix of bulk RNA-seq expression. `rownames` are bulk sample IDs, while `colnames` are gene names/IDs.

`sc.dat`: The cell-by-gene raw count matrix of bulk RNA-seq expression. `rownames` are bulk cell IDs, while `colnames` are gene names/IDs.

`cell.type.labels` is a character vector of the same length as `nrow(sc.dat)` to denote the cell type of each cell in the reference.

`cell.state.labels` is a character vector of the same length as `nrow(sc.dat)` to denote the cell state of each cell in the reference. In our example, cell states of malignant cells were obtained by sub-clustering the malignant cells from each patient, and cell states of myeloid cells were obtained by clustering myeloid cells from all patients. We define multiple cell states for these two cell types, as they contain substantial heterogeneity while also having sufficient number of cells for sub-clustering.

Please make sure that all cell states contain a reasonable number of cells, e.g. >20 or >50, so that their profile can be represented accurately.

```{r, eval = FALSE}
PBMCbulk <- read.table(file = "PBMC_ARCHIVE/Fig2b-WholeBlood_RNAseq.txt", header = TRUE,sep = "\t")
rownames(PBMCbulk) <- PBMCbulk[,1]
PBMCbulk<-PBMCbulk[,-1]

PBMCref <- read.table(file = "PBMC_ARCHIVE/Fig2ab-NSCLC_PBMCs_scRNAseq_refsample.txt", header = TRUE,sep = "\t")
rownames(PBMCref) <- PBMCref[,1]
PBMCref<-PBMCref[,-1]
head(PBMCref[,1:3])
head(PBMCbulk[,1:3])


bk.dat <- t(as.matrix(PBMCbulk,byrow = TRUE,
                      dimnames = list(rownames(PBMCbulk),
                                      colnames(PBMCbulk))))
#head(bk.dat[,1:5])
head(bk.dat[,1:5])
head(PBMCref[,1:5])
sc.dat <- t(as.matrix(PBMCref,byrow = TRUE,
                      dimnames = list(rownames(PBMCref),
                                      colnames(PBMCref))))
head(PBMCref[,1:5])


PBMCref_cell <- colnames(PBMCref)
# Cell types
PBMCcell_types <- c("B.cells", "T.cells.CD4", "T.cells.CD8", "NKT.cells", "NK.cells", "Monocytes")

library(stringr)
# Extract cell type and additional information using stringr
matches <- str_match(PBMCref_cell, paste0("(", paste(PBMCcell_types, collapse = "|"), ")\\.?(.*)"))
cell.type.labels<-matches[, 2]
cell.state.labels<-gsub("\\.", "-", matches[, 1])

sort(table(cell.type.labels))
sort(table(cell.state.labels))
table(cbind.data.frame(cell.state.labels, cell.type.labels))
```



## Filter outlier genes

Gene expressed at high magnitude, such as ribosomal protein genes and mitochondrial genes, may dominate the distribution and bias the inference. These genes are often not informative in distinguishing cell types and can be a source of large spurious variance. As a result, they can be detrimental to deconvolution. We recommend the removal of these genes.

Users may visualize the distribution of outlier genes from scRNA-seq reference. We compute the mean expression and of each gene across all cell types, and their cell type specificity scores.

Visualize and determine outlier genes from scRNA-seq data

Similarly, we may also visualize outlier genes from bulk RNA-seq. We compute the mean expression and of each gene across all cell types. As we do not have cell type level information from bulk data, we compute cell type specificity score from scRNA-seq, same as above.

Visualize outlier genes in bulk RNA-seq

```{r, eval = FALSE}
sc.stat <- plot.scRNA.outlier(
      input=sc.dat, #make sure the colnames are gene symbol or ENSMEBL ID
      cell.type.labels=cell.type.labels,
      species="hs", #currently only human(hs) and mouse(mm) annotations are supported
      return.raw=TRUE #return the data used for plotting.
      #pdf.prefix="gbm.sc.stat" specify pdf.prefix if need to output to pdf
)
bk.stat <- plot.bulk.outlier(
      bulk.input=bk.dat,#make sure the colnames are gene symbol or ENSMEBL ID
      sc.input=sc.dat, #make sure the colnames are gene symbol or ENSMEBL ID
      cell.type.labels=cell.type.labels,
      species="hs", #currently only human(hs) and mouse(mm) annotations are supported
      return.raw=TRUE
      #pdf.prefix="gbm.bk.stat" specify pdf.prefix if need to output to pdf
)
sc.dat.filtered <- cleanup.genes (input=sc.dat,
                                  input.type="count.matrix",
                                  species="hs",
                                  gene.group=c( "Rb","Mrp","other_Rb","chrM","MALAT1","chrX","chrY") ,
                                  exp.cells=5)
dim(sc.dat.filtered)

sc.dat.filtered.pc <-  select.gene.type (sc.dat.filtered,
                                         gene.type = "protein_coding")
```




## Construct a prism object.
```{r, eval = FALSE}
myPrism <- new.prism(
  reference=sc.dat.filtered.pc, 
  mixture=bk.dat,
  input.type="count.matrix", 
  cell.type.labels = cell.type.labels, 
  cell.state.labels = cell.state.labels,
  key="tumor",
  outlier.cut=0.01,
  outlier.fraction=0.1,
)
```



## Run BayesPrism

Parameters to control Gibbs sampling and optimization can be specified using gibbs.control and opt.control. Do ?run.prism for details. We recommend the use of default parameters.
```{r, eval = FALSE}
bp.res <- run.prism(prism = myPrism, n.cores=50)
```

## Report the summary statistics and Extract results
```{r, eval = FALSE}
bp.res
slotNames(bp.res)
```


"prism" is the input prism.
"posterior.initial.cellState" is the result of step 2.
"posterior.initial.cellType" is the result of step 3.
"reference.update" is the updated reference ψ
.
"posterior.theta_f" is the result of step 4.
"control_param" contains the parameters to run BayesPrism.

```{r, eval = FALSE}
# extract posterior mean of cell type fraction theta
theta <- get.fraction (bp=bp.res,
                       which.theta="final",
                       state.or.type="type")

head(theta)
# # extract coefficient of variation (CV) of cell type fraction
# theta.cv <- bp.res@posterior.theta_f@theta.cv
# 
# head(theta.cv)

saveRDS(theta_PBMC, "PBMC_ARCHIVED/PBMC_theta.rds")
#saveRDS(theta.cv_PBMC, "PBMC_ARCHIVED/PBMC_theta.cv.rds")
```



## Analysis PBMC
recall
```{r PBMC input}
B.PBMC<-read.table(file = "PBMC_ARCHIVE/Fig2b-WholeBlood_RNAseq.txt", header = TRUE,sep = "\t")
rownames(B.PBMC)<-B.PBMC[,1]
B.PBMC<-B.PBMC[,-1]
  
A.PBMC<-read.table(file = "PBMC_ARCHIVE/Fig2ab-NSCLC_PBMCs_scRNAseq_sigmatrix.txt", header = TRUE,sep = "\t")
rownames(A.PBMC)<-A.PBMC[,1]
A.PBMC<-A.PBMC[,-1]

ref.PBMC<-read.table(file = "PBMC_ARCHIVE/Fig2ab-NSCLC_PBMCs_scRNAseq_refsample.txt", header = TRUE,sep = "\t")
rownames(ref.PBMC)<-ref.PBMC[,1]
ref.PBMC<-ref.PBMC[,-1]


cyb_P.PBMC<-read.table(file = "PBMC_ARCHIVE/CIBERSORTx_PBMC_Adjusted.txt", header = TRUE,sep = "\t")
rownames(cyb_P.PBMC)<-cyb_P.PBMC[,1]
cyb_P.PBMC<-cyb_P.PBMC[,-1]

bp_P.PBMC<-readRDS("PBMC_ARCHIVE/PBMC_theta.rds")

true_P.PBMC<-read.table(file = "PBMC_ARCHIVE/ground_truth_whole_blood.txt", header = TRUE,sep = "\t")
rownames(true_P.PBMC)<-true_P.PBMC[,1]
true_P.PBMC<-true_P.PBMC[,-1]


head(A.PBMC[,1:5])
head(B.PBMC[,1:5])
head(ref.PBMC[,1:5])

head(cyb_P.PBMC)
head(bp_P.PBMC)
head(true_P.PBMC)
```

```{r PBMC transform mat}
mat.A.PBMC<-as.matrix(A.PBMC)
celltype.PBMC <- colnames(A.PBMC)
celltype.PBMC %in% colnames(true_P.PBMC)
# [1]  TRUE  TRUE  TRUE FALSE  TRUE  TRUE
# discard "NKT.cells"
mat.A.PBMC <- mat.A.PBMC[,-4]
celltype.PBMC <- celltype.PBMC[-4]


mat.true_P.PBMC<-as.matrix(true_P.PBMC[,celltype.PBMC])
mat.cyb_P.PBMC <-as.matrix(cyb_P.PBMC[,celltype.PBMC])
mat.bp_P.PBMC  <-as.matrix(bp_P.PBMC[,celltype.PBMC])

hat_true_B.PBMC <- mat.A.PBMC %*% t(mat.true_P.PBMC)
hat_cyb_B.PBMC  <- mat.A.PBMC %*% t(mat.cyb_P.PBMC)
hat_bp_B.PBMC   <- mat.A.PBMC %*% t(mat.bp_P.PBMC)

# find pair of samples in real and estimated bulk
sample.PBMC<- intersect(row.names(hat_true_B.PBMC),row.names(B.PBMC))
hat_true_B.PBMC2<-hat_true_B.PBMC[sample.PBMC,]
hat_cyb_B.PBMC2 <-hat_cyb_B.PBMC[sample.PBMC,]
hat_pb_B.PBMC2  <-hat_bp_B.PBMC[sample.PBMC,]
```

```{r performance evaluation}
library(ggplot2)
library(RColorBrewer)
library(gridExtra)
# library(knitr)

# 5.1 Cell Type Proportions Analysis
celleval<-function(hat_P,hat2_P,P){ # hat1 for CybersortX hat2 for BayesPrism
  # cell-type-level analysis
  MSE_cell<-apply((P-hat_P)^2,2,mean)
  r_cell  <-apply((hat_P-apply(hat_P,2,mean))*(P-apply(P,2,mean)),2,sum)/
    (sqrt(apply((hat_P-apply(hat_P,2,mean))^2,2,sum))*sqrt(apply((P-apply(P,2,mean))^2,2,sum)))
  # sample-level analysis
  MSE_samp<-apply((P-hat_P)^2,1,mean)
  r_samp  <-apply((hat_P-apply(hat_P,1,mean))*(P-apply(P,1,mean)),1,sum)/
    (sqrt(apply((hat_P-apply(hat_P,1,mean))^2,1,sum))*sqrt(apply((P-apply(P,1,mean))^2,1,sum)))
  # overall
  MSE_all<-mean((P-hat_P)^2)
  r_all  <-sum(hat_P-mean(hat_P)*(P-mean(P)))/
    (norm(hat_P-mean(hat_P))*norm(P-mean(P)))
  ##############################################
  # cell-type-level analysis
  MSE_cell2<-apply((P-hat2_P)^2,2,mean)
  r_cell2  <-apply((hat2_P-apply(hat2_P,2,mean))*(P-apply(P,2,mean)),2,sum)/
        (sqrt(apply((hat2_P-apply(hat2_P,2,mean))^2,2,sum))*sqrt(apply((P-apply(P,2,mean))^2,2,sum)))
  # sample-level analysis
  MSE_samp2<-apply((P-hat2_P)^2,1,mean)
  r_samp2  <-apply((hat2_P-apply(hat2_P,1,mean))*(P-apply(P,1,mean)),1,sum)/
        (sqrt(apply((hat2_P-apply(hat2_P,1,mean))^2,1,sum))*sqrt(apply((P-apply(P,1,mean))^2,1,sum)))
  # overall
  MSE_all2<-mean((P-hat2_P)^2)
  r_all2  <-sum(hat2_P-mean(hat2_P)*(P-mean(P)))/
        (norm(hat2_P-mean(hat2_P))*norm(P-mean(P)))
  
  mlist<-list(cell_level=matrix(rbind(MSE_cell,MSE_cell2,r_cell,r_cell2),nrow=4,byrow=T,
                                dimnames=
                                      list(c("MSE_cell_CybersortX","MSE_cell_BayesPrism","r_Cell_CybersortX","r_Cell_BayesPrism"),
                                           names(MSE_cell))),
              samp_level=matrix(rbind(MSE_samp,MSE_samp2,r_samp,r_samp2),nrow=4,byrow=T,
                                dimnames=
                                      list(c("MSE_samp_CybersortX","MSE_samp_BayesPrism","r_samp_CybersortX","r_samp_BayesPrism"),
                                           names(MSE_samp))),
              all_level=matrix(c(MSE_all,MSE_all2,r_all,r_all2),ncol=2,byrow=T,
                               dimnames=list(c("overall_CybersortX","overall_BayesPrism"),c("MSE_all","r_all"))))
  return(mlist)

}

# 5.2 Bulk Analysis
visualp<-function(hat_P,hat2_P,P){
df <- data.frame(x = c(P),
                 y = c(hat_P),
                 col = rep(colnames(P), each = nrow(P))
)

df2 <- data.frame(x = c(P),
                  y = c(hat2_P),
                  col = rep(colnames(P), each = nrow(P))
)
    
# Define colors dynamically based on column names
custom_colors <- brewer.pal(9, "Set1")[1:dim(P)[2]]
color_mapping_df <- setNames(custom_colors, as.character(colnames(P)))
color_mapping_df2 <- setNames(custom_colors, as.character(colnames(P)))

# Plot
p1 <- ggplot(df, aes(x = x, y = y, color = as.factor(col))) +
  geom_point(size = 4) +
  scale_color_manual(values = color_mapping_df) +
  labs(color = "Cell", x = "P", y = expression(hat(P))) +  # LaTeX expression for \hat{P}
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Diagonal line
  ggtitle("CybersortX") +
  theme_minimal()

p2 <- ggplot(df2, aes(x = x, y = y, color = as.factor(col))) +
  geom_point(size = 4) +
  scale_color_manual(values = color_mapping_df2) +
  labs(color = "Cell", x = "P", y = expression(hat(P))) +  # LaTeX expression for \hat{P}
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Diagonal line
  ggtitle("BayesPrism") +
  theme_minimal()

# Combine plots side by side
grid.arrange(p1, p2, ncol = 2)
}

```

```{r analysis result for PBMC, fig.width=12,fig.height=5}
# 5.1
table_P.PBMC<-(celleval(hat2_P=mat.cyb_P.PBMC,hat_P=mat.bp_P.PBMC,P=mat.true_P.PBMC))
table_B.PBMC<-(celleval(hat2_P=t(hat_cyb_B.PBMC2),hat_P=t(hat_pb_B.PBMC2),P=t(hat_true_B.PBMC2)))
# 5.2
visualp(hat_P=mat.cyb_P.PBMC,hat2_P=mat.bp_P.PBMC,P=mat.true_P.PBMC)

```
Tables for fraction estimation (PBMC):
      
```{r t1,echo=FALSE}
i=1
knitr::kable(table_P.PBMC[[i]], col.names = gsub("[.]", " ",  colnames(table_P.PBMC[[i]])))
```

```{r t2,echo=FALSE}
i=2
knitr::kable(t(table_P.PBMC[[i]]), col.names = gsub("[.]", " ",  rownames(table_P.PBMC[[i]])))
```

```{r t3,echo=FALSE}
i=3
knitr::kable(table_P.PBMC[[i]], col.names = gsub("[.]", " ",  colnames(table_P.PBMC[[i]])))
```

Tables for bulk estimation (PBMC):
      
```{r t2b,echo=FALSE}
i=2
knitr::kable(t(table_B.PBMC[[i]]), col.names = gsub("[.]", " ",  rownames(table_B.PBMC[[i]])))
```

```{r t3b,echo=FALSE}
i=3
knitr::kable(table_B.PBMC[[i]], col.names = gsub("[.]", " ",  colnames(table_B.PBMC[[i]])))
```



## Analysis pancreas
recall
```{r pancreas input}
B.pancreas<-read.table(file = "pancreas_ARCHIVE/pancreas_bulk_v2.txt", header = TRUE,sep = "\t")
rownames(B.pancreas)<-B.pancreas[,1]
B.pancreas<-B.pancreas[,-1]

A.pancreas<-read.table(file = "pancreas_ARCHIVE/pancreas_inferred_phenoclasses.txt", header = TRUE,sep = "\t")
rownames(A.pancreas)<-A.pancreas[,1]
A.pancreas<-A.pancreas[,-1]
dim(A.pancreas)

ref.pancreas<-read.table(file = "pancreas_ARCHIVE/pancreas_scref_v2.txt", header = TRUE,sep = "\t")
rownames(ref.pancreas)<-ref.pancreas[,1]
ref.pancreas<-ref.pancreas[,-1]


cyb_P.pancreas<-read.table(file = "pancreas_ARCHIVE/CIBERSORTx_pancreas_Adjusted.txt", header = TRUE,sep = "\t")
rownames(cyb_P.pancreas)<-cyb_P.pancreas[,1]
cyb_P.pancreas<-cyb_P.pancreas[,-1]

bp_P.pancreas<-readRDS("pancreas_ARCHIVE/pancreas_theta.rds")[,c("acinar", "alpha", "beta", "delta", "ductal", "endothelial", "epsilon", "gamma", "mast", "PSC")] # bp.posterior.initial.cellType.theta.rds") # 

true_P.pancreas<-read.table(file = "pancreas_ARCHIVE/pancreas_truth.txt", header = TRUE,sep = "\t")


head(A.pancreas[,1:5])
head(B.pancreas[,1:5])
head(ref.pancreas[,1:5])

head(cyb_P.pancreas)
head(bp_P.pancreas)
head(true_P.pancreas)
```

```{r pancreas transform mat}
mat.A.pancreas<-as.matrix(A.pancreas)
celltype.pancreas <- colnames(A.pancreas)
celltype.pancreas %in% colnames(true_P.pancreas)
# [1] TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
celltype.pancreas %in% colnames(cyb_P.pancreas)
# [1]  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE FALSE
celltype.pancreas %in% colnames(bp_P.pancreas)
# [1] TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
# discard "mast" (cyb_P)
mat.A.pancreas <- mat.A.pancreas[,-9]
celltype.pancreas <- celltype.pancreas[-9]


mat.true_P.pancreas<-as.matrix(true_P.pancreas[,celltype.pancreas])
mat.cyb_P.pancreas <-as.matrix(cyb_P.pancreas[,celltype.pancreas])
mat.bp_P.pancreas  <-as.matrix(bp_P.pancreas[,celltype.pancreas])

hat_true_B.pancreas <- mat.A.pancreas %*% t(mat.true_P.pancreas)
hat_cyb_B.pancreas  <- mat.A.pancreas %*% t(mat.cyb_P.pancreas)
hat_bp_B.pancreas   <- mat.A.pancreas %*% t(mat.bp_P.pancreas)

# find pair of samples in real and estimated bulk
sample.pancreas<- intersect(row.names(hat_true_B.pancreas),row.names(B.pancreas))
hat_true_B.pancreas2<-hat_true_B.pancreas[sample.pancreas,]
hat_cyb_B.pancreas2 <-hat_cyb_B.pancreas[sample.pancreas,]
hat_pb_B.pancreas2  <-hat_bp_B.pancreas[sample.pancreas,]
```

```{r analysis result for pancreas, fig.width=12,fig.height=5}
# 5.1
table_P.pancreas<-(celleval(hat2_P=mat.cyb_P.pancreas,hat_P=mat.bp_P.pancreas,P=mat.true_P.pancreas))
table_B.pancreas<-(celleval(hat2_P=t(hat_cyb_B.pancreas2),hat_P=t(hat_pb_B.pancreas2),P=t(hat_true_B.pancreas2)))
# 5.2
visualp(hat_P=mat.cyb_P.pancreas,hat2_P=mat.bp_P.pancreas,P=mat.true_P.pancreas)

```
Tables for fraction estimation (pancreas):

```{r t1.2,echo=FALSE}
i=1
knitr::kable(table_P.pancreas[[i]], col.names = gsub("[.]", " ",  colnames(table_P.pancreas[[i]])))
```

```{r t2.2,echo=FALSE}
i=2
knitr::kable(t(table_P.pancreas[[i]]), col.names = gsub("[.]", " ",  rownames(table_P.pancreas[[i]])))
```

```{r t3.2,echo=FALSE}
i=3
knitr::kable(table_P.pancreas[[i]], col.names = gsub("[.]", " ",  colnames(table_P.pancreas[[i]])))
```

Tables for bulk estimation (pancreas):
      
```{r t2.2b,echo=FALSE}
i=2
knitr::kable(t(table_B.pancreas[[i]]), col.names = gsub("[.]", " ",  rownames(table_B.pancreas[[i]])))
```

```{r t3.2b,echo=FALSE}
i=3
knitr::kable(table_B.pancreas[[i]], col.names = gsub("[.]", " ",  colnames(table_B.pancreas[[i]])))
```

